---
title: "HW3-YS"
author: "Yonghan Shi"
date: "3/4/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	prompt = TRUE
)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dummies)
setwd("/Users/whiz/Documents/GitHub/Econ613/Assignments/A3")
rm(list = ls())
```


```{r}
# read data
datstu <- read.csv("Data/datstu_v2.csv")
datjss <- read.csv("Data/datjss.csv")
datsss <- read.csv("Data/datsss.csv")
```

## Exercise 1 Basic Statistics
• Number of students, schools, programs

```{r}
nstu <- as.numeric(count(datstu))

#number of schools being applied by students (not that in the data of schools, not that with students admitted

nsch <- datstu %>%
  gather(key = schoolcode, value = Value, schoolcode1, schoolcode2, schoolcode3, schoolcode4, schoolcode5, schoolcode6)%>%
  summarise(distinct_school = n_distinct(Value, na.rm =TRUE))

nsch <- as.numeric(nsch)

npgm <- datstu %>%
  gather(key = choicepgm, value = Value, choicepgm1, choicepgm2, choicepgm3, choicepgm4, choicepgm5, choicepgm6)%>%
  summarise(distinct_program = n_distinct(Value, na.rm =TRUE))

npgm <- as.numeric(npgm)

table11 <- cbind(nstu, nsch, npgm)
colnames(table11) <- c("student", "school", "program")
kable(table11)
```

Note the number of school means the number of schools being applied by students (not that in the data of schools or that with students admitted).

• Number of choices (school, program)
(Hint: Create a matrix of school, programs. Convert data from Wide to Long)

```{r}
nchoice_pgm <- datstu %>%
  select(V1, choicepgm1, choicepgm2, choicepgm3, choicepgm4, choicepgm5, choicepgm6) %>%
  gather(key = choicepgm, value = program, choicepgm1, choicepgm2, choicepgm3, choicepgm4, choicepgm5, choicepgm6) 

nchoice_pgm <- nchoice_pgm %>%
  mutate(id = 1:nrow(nchoice_pgm))

nchoice_sch <- datstu %>%
  select(V1, schoolcode1, schoolcode2, schoolcode3, schoolcode4, schoolcode5, schoolcode6, jssdistrict) %>%
  gather(key = schoolcode, value = school, schoolcode1, schoolcode2, schoolcode3, schoolcode4, schoolcode5, schoolcode6)

nchoice_sch <- nchoice_sch %>%
  mutate(id = 1:nrow(nchoice_sch))

nchoice <- nchoice_sch %>%
  full_join(nchoice_pgm, by= c("id", "V1")) %>%
  select(V1, school, program, jssdistrict, choicepgm) %>%
  filter(program != "") %>%
  filter(!is.na(school))

rm(nchoice_pgm, nchoice_sch)

nch <- as.numeric(summarise(nchoice, distinct_choice = n_distinct(program, school)))
```

The number of choices is `r nch`.

• Number of students applying to at least one senior high schools in the same district to home (Suppose students live in the same district to their junior high schools)

```{r}
dat13 <- datstu %>%
  select(V1, schoolcode1, schoolcode2, schoolcode3, schoolcode4, schoolcode5, schoolcode6, jssdistrict)

datsss13 <- datsss %>%
  filter(!duplicated(schoolcode)) %>%
  dplyr::select(schoolcode, sssdistrict)

out13 <- dat13$V1
for (i in 1:6){
  dat <- dat13[,c(1,i+1,8)]
  colnames(dat) <- c("id", "schoolcode", "jssdistrict")
  dat <- dat %>% 
    left_join(datsss13, by = "schoolcode") %>%
    mutate(same = sssdistrict == jssdistrict) %>%
    select(same)
  out13 <- cbind(out13, dat)
}

colnames(out13) <- c("id","same1","same2", "same3", "same4", "same5", "same6")

out13[is.na(out13)] <- FALSE

out13 <- out13 %>% 
  mutate(atleast = same1 + same2 + same3 + same4 + same5 + same6 != 0)

rm(dat)
rm(datsss13)
rm(dat13)
nsame <- as.numeric(sum(out13$atleast))
```

The number of students applying to at least one senior high schools in the same district to home is `r nsame`.

• Number of students each senior high school admitted

```{r}
dat14 <- datstu %>%
  select(V1, score, schoolcode1, schoolcode2, schoolcode3, schoolcode4, schoolcode5, schoolcode6, choicepgm1, choicepgm2, choicepgm3, choicepgm4, choicepgm5, choicepgm6, rankplace) %>%
  filter(!is.na(rankplace))

dat14 <- dat14 %>% 
  mutate(admitted = ifelse(rankplace ==1 , schoolcode1,
                                         ifelse(rankplace ==2, schoolcode2,
                                                ifelse(rankplace==3, schoolcode3,
                                                       ifelse(rankplace==4, schoolcode4,
                                                              ifelse(rankplace==5, schoolcode5,
                                                                     ifelse(rankplace==6, schoolcode6, NA))))))) %>%
  mutate(program = ifelse(rankplace ==1 , choicepgm1,
                                         ifelse(rankplace ==2, choicepgm2,
                                                ifelse(rankplace==3, choicepgm3,
                                                       ifelse(rankplace==4, choicepgm4,
                                                              ifelse(rankplace==5, choicepgm5,
                                                                     ifelse(rankplace==6, choicepgm6, NA))))))) %>%
  select(V1, score, admitted, program)

table14 <- dat14 %>%
  group_by(admitted) %>%
  mutate(nadmit = n()) %>%
  ungroup() %>%
  filter(!is.na(admitted))%>%
  select(admitted, nadmit) %>%
  distinct() %>%
  arrange(admitted)

colnames(table14) <- c("schoolcode","size")

kable(head(table14))
```

• The cutoff of senior high schools (the lowest score to be admitted)

```{r}
table15 <- dat14 %>%
  group_by(admitted) %>%
  mutate(cutoff = min(score)) %>%
  ungroup() %>%
  filter(!is.na(admitted))%>%
  select(admitted, cutoff) %>%
  distinct()%>%
  arrange(admitted)

colnames(table15) <- c("schoolcode","cutoff")

kable(head(table15))
```

• The quality of senior high schools (the average score of students admitted)

```{r}
table16 <- dat14 %>%
  group_by(admitted) %>%
  mutate(quality = mean(score))%>%
  ungroup() %>%
  filter(!is.na(admitted))%>%
  select(admitted, quality) %>%
  distinct()%>%
  arrange(admitted)

colnames(table16) <- c("schoolcode","quality")

kable(head(table16))
```

## Exercise 2 Data

```{r}
datsss2 <- datsss %>%
  filter(!duplicated(schoolcode)) %>%
  select(!V1)

dat2 <- dat14 %>%
  group_by(admitted, program) %>%
  mutate(size = n()) %>%
  mutate(cutoff = min(score)) %>%
  mutate(quality = mean(score))%>%
  ungroup() %>%
  filter(!is.na(admitted))%>%
  select(admitted, program, size, cutoff, quality) %>%
  distinct()

colnames(dat2) <- c("schoolcode" , "program", "size", "cutoff", "quality")

dat2 <- dat2 %>%
  left_join(datsss2, by = "schoolcode")%>%
  arrange(schoolcode)

kable(head(dat2))
```

## Exercise 3 Distance

```{r}
colnames(nchoice) <- c("V1", "schoolcode", "program", "jssdistrict", "choice")

datsss3 <- datsss2 %>%
  select(schoolcode, ssslong, ssslat)

dat3 <- nchoice %>%
  left_join(datjss, by = "jssdistrict") %>%
  left_join(datsss3, by = "schoolcode") %>%
  mutate(choice_rank = ifelse(choice == "choicepgm1", 1,
                                         ifelse(choice == "choicepgm2", 2,
                                                ifelse(choice == "choicepgm3", 3,
                                                       ifelse(choice == "choicepgm4", 4,
                                                              ifelse(choice == "choicepgm5", 5, 6
                                                                     ))))))  %>%
  select(V1, schoolcode, program, point_x, point_y, ssslong, ssslat, choice_rank)

colnames(dat3) <- c("V1", "schoolcode", "program", "jsslong", "jsslat", "ssslong", "ssslat", "choice_rank")

dat3 <- dat3 %>%
  mutate(dist = sqrt((69.172*(ssslong-jsslong)*cos(jsslat/57.3))^2 + (69.172*(ssslat-jsslat))^2))

kable(head(dat3[,c(1:3,8,9)]))
```

## Exercise 4 Dimensionality Reduction


• Recode the schoolcode into its first three digits (substr). Call this new variable scode_rev.

• Recode the program variable into 4 categories: arts (general arts and visual arts), economics (business and home economics), science (general science) and others. Call this new variable pgm_rev.

• Create a new choice variable choice_rev.

• Recalculate the cutoff and the quality for each recoded choice.

• Consider the 20,000 highest score students.

• The rest of the assignment uses the recoded choices and the 20,000 highest score students.

```{r}
datind<- datstu %>%
  select(V1, score, rankplace)

dat4 <- nchoice %>%
  #Recode the schoolcode
  mutate(scode_rev = substr(schoolcode, 1, 3)) %>%
  #Recode the program
  mutate(pgm_rev = ifelse(program == "General Arts" | program == "Visual Arts", "Arts",
                                         ifelse(program == "Business" | program == "Home Economics", "Economics",
                                                ifelse(program == "General Science", "Science", "Others"))))%>%  
  #Create a new choice variable
  unite("choice_rev", scode_rev, pgm_rev, sep = " ", remove= FALSE) %>%
  mutate(choice_rank = ifelse(choice == "choicepgm1", 1,
                                         ifelse(choice == "choicepgm2", 2,
                                                ifelse(choice == "choicepgm3", 3,
                                                       ifelse(choice == "choicepgm4", 4,
                                                              ifelse(choice == "choicepgm5", 5, 6
                                                                     )))))) %>%
  left_join(datind, by = "V1") %>%
  filter(!is.na(rankplace)) %>%
  mutate(admitted = rankplace == choice_rank) %>%
  select(V1, scode_rev, pgm_rev, choice_rev, choice_rank, score, admitted)

#calculate cutoff and quality
dat4choice <- dat4 %>%
  filter(admitted == TRUE) %>%
  group_by(choice_rev) %>%
  #Recalculate the cutoff and the quality
  mutate(cutoff = min(score)) %>%
  mutate(quality = mean(score))%>%
  ungroup() %>%
  distinct(choice_rev, cutoff, quality)

# Add distance factor
datdist <- select(dat3, V1, dist, choice_rank)

dat4 <- dat4 %>%
  left_join(dat4choice, by = "choice_rev") %>%
  left_join(datdist, by = c("V1","choice_rank"))

# Select 20000 with the highest score
stu20000 <- datstu %>%
  select(V1, score, agey, male) %>%
  arrange(desc(score))

stu20000 <- stu20000[1:20000,]

dat4 <- stu20000 %>%
  left_join(dat4, by = c("V1" , "score"))

kable(head(dat4))
```

## Exercise 5 First Model

• Propose a model specification. Write the Likelihood function.
```{r}
## interested in the first choice
dat5 <- dat4 %>% 
  filter(choice_rank == 1) %>%
  mutate(choice=as.numeric(factor(choice_rev,ordered=TRUE)))
```

Test score is the same for each students, so we use multinomial logit model. 

```{r echo=TRUE}
## X and Y for the model
X1 <- dat5 %>%
  mutate(intercept = 1) %>%
  select(intercept, score)

X1 <- as.matrix(X1)

Y1 <- dummy("choice_rev", data = dat5, sep="_")

beta1 = solve(t(X1) %*% X1) %*% t(X1) %*% Y1

# Likelihood Function for multinomial logit model
mlogit.ll <- function(beta) {
  B <- matrix(beta, nrow = 2, byrow = TRUE)
  B[,1] <- 0
  est <- exp(X1 %*% B)
  Pr <- t(apply(est, 1, function(x) x / sum(x)))
  Pr <- Pr*Y1
  loglikelihood <- sum(log(rowSums(Pr)))
  return(-loglikelihood)
}
```

• Estimate parameters and compute marginal effect of the proposed model.
```{r message=FALSE, warning=FALSE}
# Optimization

#res <- optim(beta1, mlogit.ll, method = "BFGS")
#fit1 <- nlm(mlogit.ll, beta1)
#out5 <- matrix(fit1$estimate, nrow = 2, byrow = TRUE)

library(nnet)
fit1 <- multinom(choice_rev ~ score, dat5)

out5 <- t(coef(fit1))


# Marginal Effect
marginal.mlogit <- function(beta){
  est <- exp(X1 %*% beta)
  Pr <- t(apply(est, 1, function(x) x / sum(x)))
  B1 <- matrix(rep(out5[1,],each=nrow(Pr)),nrow=nrow(Pr))
  B2 <- Pr %*% t(beta)
  B2 <- matrix(rep(B2[,1],each=ncol(beta)), ncol=ncol(beta), byrow=TRUE)
  dB <- B1 - B2
  PrOut <- Pr*dB
  ME <- data.frame(t(apply(PrOut, 2, mean)))
  return(ME)
}

ME.mlogit <- marginal.mlogit(out5)
```

Parameters:
```{r}
kable(out5[,1:6])
```

Marginal Effect:
```{r}
kable(ME.mlogit[,1:6])
```

## Exercise 6

• Propose a model specification. Write the Likelihood function.

School Quality is the same for each choice, so we use conditional logit model.

```{r echo=TRUE}
## X and Y for the model

X2 <- dat5%>%
  select(choice_rev, quality) %>%
  filter(!duplicated(choice_rev)) %>%
  arrange(choice_rev)
  
X2 <- t(data.frame(X2, row.names=1))

X2f <- as.matrix(X2-X2[1])

Y2 <- dummy("choice_rev", data = dat5, sep="_")

beta2 <- runif(246,-1,1)

# Likelihood Function for conditional logit model
clogit.ll <- function(beta) {
  B <- beta2[1]
  A <- as.matrix(beta2[2:246])
  A[1] <- 0
  est <- exp(X2f * B + ( matrix(1,nrow = nrow(X2f * B),ncol = 1) %*% t(A)))
  Pr <- t(apply(est, 1, function(x) x / sum(x)))
  Pr <- matrix(rep(Pr,each=nrow(Y2)),nrow=nrow(Y2))
  Pr <- Pr * Y2
  loglikelihood <- sum(log(rowSums(Pr)))
  return(-loglikelihood)
}

```

• Estimate parameters and compute marginal effect of the proposed model.

```{r}
# Optimization
fit2 <- nlm(clogit.ll, beta2)

out6 <- as.data.frame(t(fit2$estimate[c(1,3:246)]))

colnames(out6) <- c("intercept", colnames(Y2[,2:245]))

# Marginal Effect
marginal.clogit <- function(beta){
  beta <- as.matrix(beta)
  B <- beta[1]
  est <- X2* B
  A <- c(0,beta[2:245])
  est <- exp(est + ( matrix(1,nrow = nrow(est),ncol = 1) %*% t(A)))
  Pr <- t(apply(est, 1, function(x) x / sum(x)))
  BPr <- B * colSums(Pr) 
  BPr_diag <- matrix(rep(BPr,each=245),nrow=245) * diag(245)
  BPr_sq <- B * (t(Pr) %*% Pr)
  ME <- (BPr_diag - BPr_sq)/nrow(X2)
  return(ME)
}

ME.clogit <- marginal.clogit(out6)
```

Parameters:
```{r}
kable(out6[1:6])
```

Marginal Effect:
```{r}
ME.clogit[1:6,1:6]
```
## Exercise 7 Counterfactual Simulations

• Explain and justify, which model (first or second model) you think is appropriate to conduct this exercise.

I think the first model is appropriate, as for the second model, moving "others" out won't affect the result much.

• Calculate choice probabilities under the appropriate model.

```{r}
Pr72 <- as.data.frame(t(colSums(t(apply(exp(X1 %*% cbind(0, out5)), 1, function(x) x / sum(x))))/nrow(X1)))

colnames(Pr72)[1] <- "100 Arts"

kable(Pr72[,1:6])
```

• Simulate how these choice probabilities change when these choices are excluded.
```{r message=FALSE, warning=FALSE}
dat7 <- dat5 %>%
  filter(pgm_rev != "Others")

fit3 <- multinom(choice_rev ~ score, dat7)

out7 <- t(coef(fit3))

X3 <- dat7 %>%
  mutate(intercept = 1) %>%
  select(intercept, score)

X3 <- as.matrix(X3)

Pr73 <- as.data.frame(t(colSums(t(apply(exp(X3 %*% cbind(0, out7)), 1, function(x) x / sum(x))))/nrow(X3)))

colnames(Pr73)[1] <- "100 Arts"

kable(Pr73[,1:6])
```

The choice probabilities become higher when excluding those choices.


